Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-16 paper/current/main.tex
update intro and first para

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in past eras, it holds less importance to a world connected by rapid transit, complex international relations, and dense airline networks.
For instance, a person currently living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend university in Canada rather than the closer though diplomatically isolated North Korea. 
Despite its complexity, geographic distance is used as the basis for the most prominent models of mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal}.
So-called \textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} have been proposed that attempt to learn alternative notions of distance from real-world data or \textit{a priori} relationships between geographic entities.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, originally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
Neural embedding is also computationally efficient, scalable for large datasets, robust to noise, and is readily interpretable through the application of simple vector arithmetic.

We use scientific mobility, inferred from affiliations, to demonstrate the neural embedding approach. 
Scientific mobility is a central aspect of the increasingly-globalized scientific system~\autocite{czaika2018globalisation, altbach2004globalization}, being strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}.
For nations and organizations hoping to win in the global competition for talent~\autocite{box2008competition}, understanding the landscape of global scientific mobility is essential. 
However, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}.
One roadblock is the lack of a meaningful measure of the functional distance between countries, cities, and organizations.

Here, scientific mobility is operationalized using researcher's organizational affiliations derived from their publication histories indexed in a large-scale bibliographic database. 
The most fine-grained geographical unit in this data is the organization, such as a university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
These trajectories encode both asynchronous migration, and synchronous mobility---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see Supplementary Materials).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors who were affiliated with both organizations during the time period is proportional to the organization's sizes (the sum of both mobile and non-mobile authors, see Data and Methods), $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
The distance function is defined in terms of either the geographic distance between organizations or the embedding distance. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
We argue that the performance of a distance measure can be observed from its ability to explain expected mobility between organizations and to make predictions using the gravity model.

@@

The functional distance derived from the embedding explains and predicts scientific mobility better than geographic distance.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c--d), which also holds for the subsets of only domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility (Fig.~\ref{fig:supp:gravity_correlation}g--h).
The embedding distance defined from the cosine distnace between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see supporting materials) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results unequivocally argue that the distance learned by the embedding method better captures actual mobility patterns than does geographic distance. 
We speculate that this results from an underlying mathematical connection between the neural embedding architecture and the gravity model of mobility.

>>>>

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in the past, it holds less importance to a world connected by rapid transit, complex international relations, and dense airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend universities in Canada than those in North Korea. 
While geographic distance is still used as the basis for the most prominent models of mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} model, so-called \textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} have been proposed to learn alternative notions of distance from real-world data or \textit{a priori} relationships between geographic entities.  Yet, limitations of the existing functional distance measure still are not being overcome.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, originally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
Neural embedding is also computationally efficient, scalable for large datasets, robust to noise, and can extract rich relationship between the entities through the application of simple vector arithmetic.

We use scientific mobility, inferred from affiliations, to demonstrate the effectiveness of our approach
Scientific mobility is a central aspect of the increasingly-globalized scientific system~\autocite{czaika2018globalisation, altbach2004globalization}, being strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}.
For nations and organizations hoping to win in the global competition for talent~\autocite{box2008competition}, understanding the landscape of global scientific mobility is essential. 
However, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}.
Underlying these challenges is the lack of a meaningful functional distance measure between countries, cities, and organizations.

Here, scientific mobility is operationalized using researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database. 
The most fine-grained geographical unit in this data is the organization, such as a university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
These trajectories encode both asynchronous migration, and synchronous mobility---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see Supplementary Materials).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors who were affiliated with both organizations during the time period is proportional to the organization's sizes (the sum of both mobile and non-mobile authors, see Data and Methods), $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
The distance function is defined in terms of either the geographic distance between organizations or the embedding distance. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

The functional distance derived from the embedding explains and predicts scientific mobility better than geographic distance.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c--d), which also holds for the subsets of only domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility (Fig.~\ref{fig:supp:gravity_correlation}g--h).
The embedding distance defined from the cosine distnace between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see supporting materials) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results unequivocally argue that the distance learned by the embedding method better captures actual mobility patterns than does geographic distance.
